---
title: "Lab 01 - Intro to R"
description: "Week 1: An iIntroductory tutorial for programming in the R language"
output: 
  learnr::tutorial:
    fig_caption: no
    progressive: true
    allow_skip: true
    toc: true
    toc_depth: 3
    theme: readable
runtime: shiny_prerendered
---


```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.checker = gradethis::grade_learnr)
knitr::opts_chunk$set(echo = FALSE)
```




## Intro to Programming in R

Welcome to this tutorial! R is a programming language and free software 
environment for statistical computing and graphics supported by the R Core Team
and the R Foundation for Statistical Computing.

Even if you are excellent at programming, if you haven't used R before you need 
a bit of help to translate your existing knowledge to this other language. 

If you don't feel that your programming skills are very strong yet, this tutorial
will also help you to get some practice and acquire some concept that will get 
you to the next level!

So, let's get started!



## Checking installation

Before starting to use R, we need to check whether you have the *right* software 
installed. If you are running this tutorial, you must have something installed. 
However, we need to make sure you have the right versions and that you have all 
the packages we need.

### R Version 

You need to have installed R version 4.1.1 and this tutorial is going to check it
for you. Please read the message below.
```{r rcheck, include = TRUE}
# R check version (required 4.1.1)

major <- R.Version()$major
minor <- R.Version()$minor

if((major == 4) & (minor == 1.1 )) {

  cat("Great! You have the right version!")

} else {

  cat("Sorry! You have the wrong version. Please update R to version 4.4.1")

}


```


### R Studio Version

You need to have installed RStudio version 1.4.1717 or above.
Please read the message below.

```{r rstudio check, include = TRUE}
# R studio check version (required 1.4.1725 or above)
ver <- rstudioapi::versionInfo()$version

ver <- sub("1.", "", ver)

if(ver < 4.1717) {

  cat("Sorry! Your Rstudio is not update. Please download version 1.4.1725 or above")

} else {

  cat("Great! Your Rstudio is up to date!")
}


```



### Packages

You need to have a few packages installed:



```{r}
# Function to check whether package is installed

# insert here packages that the students need to have installed.
pkg <- c("SNA4DSData", "igraph", "sna", "gradethis", "papaja", "ina")


is.installed <- function(mypkg){
  is.element(mypkg, installed.packages()[,1])
}



for (i in 1:length(pkg)){

if (isFALSE(is.installed(pkg[i]))){

  install.packages(pkg[i])

} else {

  cat(paste0("Great! You have the ", pkg[i], " package\n"))

  }
}
```

Do you have `SNA4DSData` installed? If not run this line of code on your *R*
console:

`remotes::install_github("SNAnalyst/SNA4DSData", dependencies = TRUE)`

Ok, check up completed! 
Did you pass the checks? Yes? Great! Let's get started for real!
If not, please close the tutorial, update/download and come back!



## Operators, Variables and Conditions

As you probably already know, to program you need to master a handful of operators. 
Let's have an overview together!

### Arithmetic

We begin with something very simple. You are surely familiar with basic math
operations, but you might not know how to write them in the R language. 

Let's take a look

* \+  summation 
* \-  subtraction
* \*  multiplication
* /   division
* \%/%  integer division
* ^  exponentiation
* sqrt   square root
* log logarithm 

Why don't you try an operation of your choice? For instance 20 %/% 3?

```{r try_something, exercise=TRUE}


```

* Take the square root of the log of 5. Then click Submit Answer.

```{r grade_code1, exercise = TRUE}

```

```{r grade_code1-solution}
sqrt(log(5))
```

```{r grade_code1-check}
gradethis::grade_code(correct = "Excellent work.")
```



### Comparison

Other basic, but useful thing to know is how to compare to values

* == equal
* \> larger than
* < smaller than
* \>= larger or equal to
* <= smaller or equal to
* != different than 

* Can you write that the square root of 2 is diffent than 5 at the power of 3?

```{r grade_code2, exercise = TRUE}

```

```{r grade_code2-solution}
sqrt(2) != 5^3
```

```{r grade_code2-check}
gradethis::grade_code(correct = "Excellent work.")
```


### Logical

how about expressing logical statements in R?

* & and - It will return true when both conditions are true
* && and - It will return true when both conditions are true but it works only for the first element
* \| or - It will return true when at-least one of the conditions is true
* \|| or - It will return true when at-least one of the conditions is true but it works only for the first element
* ! not - If the condition is TRUE , logical not operator will return FALSE

Can you write that 5 or 10 are larger or equal to 5?

```{r grade_code3, exercise = TRUE}

```

```{r grade_code3-solution}
5 | 10 >= 5 
```

```{r grade_code3-check}
gradethis::grade_code(correct = "Brilliant!")
```


### Variable Assignment

How do we make sure that we store the work we are doing?
If you are a Pythonist, you are use to this 

C = 2 + 2 

In R, it runs too, but it is better not to do that

This is the way you should specify a variable in R

C <- 2 + 2

If you want R to print it while you specify it, you use round brackets 

(C <- 2 + 2)

Wanna give it a go? Try these three options

```{r try_something1, exercise=TRUE, echo = TRUE}


```


### Using Conditions

Operators are usually used to check conditions. These concepts are the foundation
of very advanced programming operations!

You can write

(a == b) == TRUE

but it is not very nice.

isTRUE(a == b)

It ss much better

the same in this case

(c < d) == FALSE

doesn't it look nicer? 

isFALSE(c < d)

check what are a and be below. Can you check whether 

it is TRUE that a is equal to b?
it is FALSE that a + b is different than 8?


```{r grade_code4, exercise = TRUE}
a <- 3
b <- 5



```

```{r grade_code4-solution}
a <- 3
b <- 5

isTRUE(a == b)
isFALSE((a + b) == 8)


```

```{r grade_code4-check}
gradethis::grade_code(correct = "Great job!")
```



## Make your code readable


There are three things that improve the readability of your code:

* Commenting 
* Indenting
* Spacing

### Commenting 
It is also good if you comment your code every time. So you will be able to read
it in the future. Even more so if you are doing collaborative work. In R you 
comment using #

Can you write on the top of this operation that it is a summation? 

```{r try_something2, exercise=TRUE, echo = TRUE}

2 + 2

```


### Indenting
In R indentation is important to keep your code readable, but it is not crucial. 
It will be running even if you write it in the most terrible way. R uses curly 
braces and that's what matters the most.

.center[Just please DON'T!]

R studio will automatically indent your code if you select it and click CRTL I. 
Try to write some messed up code and to indent it afterward.

```{r try_something3, exercise=TRUE, echo = TRUE}

   2 + 2

```


### Spacing 

Also, it is nice if you leave space between the characters. It will be harder to 
read a page of code with no space. Your eyes will thank you! 

Please, fix this for me... 


```{r try_something311, exercise=TRUE, echo = TRUE}

c<-(2+3+4)/4

```


## R Objects
Programming languages can be of two types: functional and object-oriented. Java
is an example of an object-oriented language, while Python is a functional one. 
One type is not better than the other. It is better for doing certain things. 

R is also a functional language; still, there are ways to specify the features 
of an object (protocols S3 and S4). You would need to learn how 
to do it if you decide to write a package in R while using R for data 
analysis, you don't need that. 

As a data analyst, you need the ability to check the object you are working with 
and to change its features if needed. 

Let's start to check the object types that we can have in R.

### Types of Variables

To make the best of the R language, you’ll need a strong understanding of the 
basic data types and how to operate on them.

* _character_: "a", "swc"
* _numeric_: 2, 15.5
* _integer_: 2L (the L tells R to store this as an integer)
* _logical_: TRUE, FALSE
* _complex_: 1+4i (complex numbers with real and imaginary parts) (we are not going to use it)
* _raw_: it is used to holds raw bytes (we are not going to use it)

Can you guess what type are the following variables? 

```{r try_something4, exercise=TRUE, echo = TRUE}

(var <- "2.5")

(var1 <- TRUE)

(var2 <- "FALSE")

(var3 <- 2L)

(var4 <- 2.5)


```

Well, you tried to guess the type of this variables by heart, but there are 
better ways to do it. 

The function `typeof()` is native to R. This means that is inbuilt and part of the
*R-base* language. In practice? You don't need to call any library 
to use it. This is the first of a long series of helpful functions in *R-base* 
that you will get to know.   

`typeof()` will able to tell you how your variable is stored in the memory of 
your computer. 

Let's try 

```{r try_something5, exercise=TRUE, echo = TRUE}

var <- "2.5"
typeof(var)

var1 <- TRUE
typeof(var1)

var2 <- "FALSE"
typeof(var2)

var3 <- 2L
typeof(var3)

var4 <- 2.5
typeof(var4)
```
What happened to var4? We have a new label we were not familiar with!

*double* what is a double? 

A double is a numeric variable. The function is telling you that it is not 
stored as an integer, hence it could have decimal points. 

If we use the function `mode()` we can make sure that a double is still a numeric
variable

```{r try_something511, exercise=TRUE, echo = TRUE}

var4 <- 2
typeof(var4)
mode(var4)

var5 <- 2L
typeof(var5)
mode(var5)


```


Be careful though, also the integers are numeric if you use the `mode()` function.


### Vectors of objects

The examples below had only one value stored in the variables, but you can store
many more and you can check the type in the same way. 

```{r try_something6, exercise=TRUE, echo = TRUE}
var <- c(0.2, 4.5, 5.3)
typeof(var)

var1 <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
typeof(var1)

var2 <- c('TRUE', 'FALSE', 'TRUE', 'TRUE', 'FALSE')
typeof(var2)

```


### Change Type

You can also change the type of your variables 

```{r try_something7, exercise=TRUE, echo = TRUE}
c<- c(0,1,1,0)
typeof(c)
c1 <- as.character(c)
typeof(c1)
c2 <- as.logical(c)
typeof(c2)

# careful though when you do these kind of transformations
(c3 <- as.character(c2))

```

*c3* is now very different than *c*. If you don't pay attention you will mess 
up your data very badly!

### class

While `typeof()` and `mode()` tell you how the variable is stored (low-level), 
another function named `class()` tells you what is the function that an object
takes in the specific programming session where you are working on (high-level).

Try to play a bit with this function

```{r try_something8, exercise=TRUE, echo = TRUE}
class(FALSE)
```
You will learn the practical differences with practice.

For the data analysis that we do it is better to use `class`.

### Factors

Factors are derived data types generally used to group variables into a fixed 
number of unique categories or levels. 

Let's use one of the most popular factor variables: gender.

Let's assume this is a group of friends that takes data science classes together. 
We have females (F), males (M), and non-binary people identified with 'O' (other).

This information is stored in the variable *a*. Let's check what type of 
variable is *a*

```{r try_something9, exercise = TRUE, echo = TRUE}
a <- c("M", "F", "O", "F", "O", "M", "F", "O")
a <- as.factor(a)

typeof(a)
mode(a)
class(a)

```

Only using class we will learn it is a factor. But a factor variable has also 
another dimension: **levels**, that we can explore with the function `levels()`.

```{r try_something10, exercise=TRUE, echo = TRUE, exercise.setup = "try_something9"}

typeof(a)
mode(a)
class(a)
levels(a)
```

**levels**, tells us the categories contained inside a factor variable. 


We can explore a factor variable in one go by using the function `attributes()`

```{r try_something11, exercise=TRUE, echo = TRUE, exercise.setup = "try_something9"}
attributes(a)
```
Why does `typeof()` report that *a* is an integer and `mode()` that it is 
numeric?

A factor has a numeric representation that corresponds to the order of the 
categories listed in *levels*

* F = 1
* M = 2
* O = 3

let's check if it's true

```{r try_something12, exercise = TRUE, echo = TRUE, exercise.setup = "try_something9"}

a
as.numeric(a)
```

it matches!

### Ordinal 

An ordinal variable is similar to a categorical variable. The difference between 
the two is that there is a clear ordering of the categories.

Since each category corresponds to a number and these numbers are ordered, 
factors can also be employed to store ordinal data.

In the examples we used for factors, there is no natural order, Female, Male, 
and Gender Other are not ranked.

Let's consider a contest for the hottest chili where ten judges have to rate 
each chili pots as

* not spicy = 1
* a bit spicy = 2
* mildly spicy = 3
* very spicy = 4
* Waw! Fire! = 5

Yes, there is an ordered ranking here. 

Let's code it up
```{r try_something13, exercise = TRUE, echo = TRUE}
grades <- c('mildly spicy', 'mildly spicy', 'very spicy', 'a bit spicy', 'Waw! Fire!', 
            'not spicy', 'very spicy', 'mildly spicy', 'a bit spicy', 'a bit spicy')


grades <- as.factor(grades)

attributes(grades)
```

Wait, the levels don't match the order! We need to fix this. We can do it with 
the `levels()` function.

```{r try_something14, exercise = TRUE, echo = TRUE, exercise.setup = "try_something13"}
levels(grades)<- c("not spicy", "a bit spicy", "mildly spicy", "very spicy", "Waw! Fire!")

attributes(grades)
```


yes! Now it is correct! Try to compare the original vector and the one turned into 
numeric to see whether they match!

```{r grade_code5, exercise = TRUE, exercise.setup = "try_something14"}



```

```{r grade_code5-solution}


grades
as.numeric(grades)


```

```{r grade_code5-check}
gradethis::grade_code(correct = "Great job!")
```


You can also specify the rank of the factors directly when you create it 
using the `factor()` function. Why don't you try to figure out how to do it? 


```{r grade_code6, exercise = TRUE, echo = TRUE}

grades <- c('mildly spicy', 'mildly spicy', 'very spicy', 'a bit spicy', 'Waw! Fire!', 
            'not spicy', 'very spicy', 'mildly spicy', 'a bit spicy', 'a bit spicy')

# levels = c("not spicy", "a bit spicy", "mildly spicy", "very spicy", "Waw! Fire!"

grades <- factor(grades)

attributes(grades)

```


```{r grade_code6-solution}


grades <- c('mildly spicy', 'mildly spicy', 'very spicy', 'a bit spicy', 'Waw! Fire!', 
            'not spicy', 'very spicy', 'mildly spicy', 'a bit spicy', 'a bit spicy')


grades <- factor(grades, levels = c("not spicy", "a bit spicy", "mildly spicy", "very spicy", "Waw! Fire!"))

attributes(grades)


```

```{r grade_code6-check}
gradethis::grade_code(correct = "Great job!")
```


## Looking for help

If you struggled with the exercise you just tried, it is just because you still 
don't know how to find help in *R*. You needed to know that *levels* is also one
of the arguments of the `factor()` function. 

R offers you a quick way to get info about its functions (and the functions of 
the packages too. More about that later)

if you type a question mark before the function name, all the information you 
need will pop up on your R studio window (or open on browser)

Uncomment the code below and give it a go!

```{r try_something15, exercise = TRUE, echo = TRUE}
# ?factor

```


## Missing Data

We discussed different types of data, but we also need to take into account that
sometimes (often), the data is missing. 

R stores missing values in two ways

* NA: (Not Available) is a missing value indicator. It suggests that a value 
should be present but is unknown. R considers NA as logical.

* NULL: implies that a value should not be present and it is not there.

See for yourself


```{r try_something16, exercise = TRUE, echo = TRUE}
a <- NA
class(a)


b <- NULL
class(b)
```

If you want to check whether a vector contains NAs, you can use `is.na()`

```{r try_something17, exercise = TRUE, echo = TRUE}
v <- c(1, 2, 3, NA, 5, 6)
is.na(v)

```

If you want to get rid of the NAs, you can use `na.omit()`

```{r try_something18, exercise = TRUE, echo = TRUE}
v <- c(1, 2, 3, NA, 5, 6)
v1 <- na.omit(v)

length(v)

length(v1)

```

Be careful though! If you omit NAs, the length of your vector will be affected. 
Keep this monitored with the `length` function. 


## Why are data types important? 

Let's say that you are a teacher and there are eight pupils in your class (boys
and girls)

You are writing an end-of-the-year report to show that there is gender equality.

We have a pupil variable, and you want to plot it so that your audience can 
appraise this information quickly.

How do we do? You can use the function `hist()` to make a histogram.

Tip: use `class()` to check the data type.


```{r grade_code7, exercise = TRUE, echo = TRUE}

p <- c("B", "B", "G", "G", "G", "G", "B", "B")


```


```{r grade_code7-solution}

p <- c("B", "B", "G", "G", "G", "G", "B", "B")

p <- as.numeric(as.factor(p))

hist(p)


```

```{r grade_code7-check}
gradethis::grade_code(correct = "Brilliant!")
```


Unless you pass to the function the class it requires, it won't work!



## Data Structures
To make the best of the R language, you’ll need a strong understanding of the 
basic data types and data structures that this language handles and how to 
operate with them. We covered data types already, let's move on to data structures.

There are 5 data structures in *R*: 

* vector 
* data frame 
* matrix
* array
* list 

### Vector

A vector, or atomic vector, is the simplest data structure available in R and we 
already used it when discussing object types. It consists of an ordered set of 
values of the _same_ type and/or class and it can be created using the function
`c()`, where _c_ means _combine_. 

Why don't you create a vector named `v` that contains numbers from 1 to 5? 
Remember that combining elements with different classes or type does not work. 


```{r try_something19, exercise = TRUE, echo = TRUE}


```

### Data Frame

A data frame is what comes closest to our perception of a traditional data table. 
Each column can host different types or classes of data. You can create one in R
with the function `data.frame()`. Every column needs to have the same length.

Try to run the code below 

```{r try_something20, exercise = TRUE, echo = TRUE}
(df <- data.frame(col1 = 'one', col2 = 1:10))

class(df)

```

In the example above it will repeat the same element for the length of the data
frame.

Can you insert the variables *a* and *b* as two columns in a data frame named
*df* and print it?


```{r grade_code71, exercise = TRUE, echo = TRUE}

a <- 1:10
b <- LETTERS[1:10]


```


```{r grade_code71-solution}

a <- 1:10
b <- LETTERS[1:10]

(df <- data.frame(col1 = a, col2 = b))


```

```{r grade_code71-check}
gradethis::grade_code(correct = "Brilliant!")
```


You can also *bind* the two colums together and insert them in the data frame. 


```{r try_something21, exercise = TRUE, echo = TRUE}
a <- 1:10
b <- LETTERS[1:10]

df <- data.frame(cbind(a, b))
colnames(df) <- c('col1', 'col2')
df
```

In this option we used three functions: 

* `data.frame()`
* `cbind()`
* `colnames()`

The latter two can be also used with rows: 

* `rbind()`, will bind rows (surprise surprise!)
* `rownames()`, will allow you to over scribe the name of the rows in your data frame.

can you make the example above work with `rbind()`? Think carefully about what 
you need to change.

```{r grade_code8, exercise = TRUE, echo = TRUE}

a <- 1:10
b <- LETTERS[1:10]


```


```{r grade_code8-solution}

a <- 1:10
b <- LETTERS[1:10]

df <- data.frame(rbind(a, b))
colnames(df) <- c('col1', 'col2', 'col3', 'col4', 'col5', 'col6', 'col7', 'col8', 'col9', 'col10')
df

```

```{r grade_code8-check}
gradethis::grade_code(correct = "Bravo! It is still colnames()!")
```



### Matrix
A matrix in R is like a data frame but it can host only one type or class of data.
It can either be numeric, character etc...
The parameters *nrow* and *ncol* define the matrix dimension. 

In this social *network class* we will make use of TONS of matrices!

Try to run this code.
```{r try_something22, exercise = TRUE, echo = TRUE}


(m <- matrix(0, nrow = 5, ncol = 6))
class(m)
```


If I tell you to use the function `as.matrix()`, can you make these two variables
into a matrix using `cbind()`?


```{r grade_code9, exercise = TRUE, echo = TRUE}

a <- LETTERS[11:20]
b <- LETTERS[1:10]

```


```{r grade_code9-solution}

a <- LETTERS[11:20]
b <- LETTERS[1:10]

m <- as.matrix(cbind(a, b))


```

```{r grade_code9-check}
gradethis::grade_code(correct = "Well done!")
```



### Array
A matrix has 2 dimensions. If you need more than two, you employ an array. It
can be 3, 4 or N dimensions as long as they store data of the same type or class.

Run the code below to create a 3x3x3 array filled out with *0s*.
```{r try_something23, exercise = TRUE, echo = TRUE}

(ar <- array(0, c(3,3,3)))

```


We are not going to use arrays very often, but still you should be familiar with
them.

### List 
You can think of a list as a more powerful kind of vector. We call it recursive
vector. Its super power consists of the fact that each element can be a different
data type, class, or even structure. 
This implies that each element of a list can hold complex objects such as 
matrices, data frames and other list objects too!

A list is constructed using the list() function. For example, the following list
consists of 3 components: 

* A: a two-column data frame 
* B: a two element logical vector 
* C: a 3x3 square matrix

```{r try_something24, exercise = TRUE, echo = TRUE}

A <- data.frame(col1 = 'one', col2 = 1:10)
B <- c(TRUE, FALSE)
C <- matrix(0, 3, 3)

(L <- list(A, B, C))

```


## Reading files in R

It is time to import real data into R! Let's take a look at how to import the
most common data extensions.


### Setting your directory

When you start to work in R (as in any other language), you need to set up your working directory. In fact, you won't be able to import any data unless R knows where to find it first. 

If you simply want to check what is the current working directory, you use the function `getwd()`. 

If you want to change the current working directory, you use `setwd()`.

The way you write the 'address' of your directory changes according to your 
operative system. While on Windows, your directory is usually located in *C*, 
Mac usually locates directories in its *home*. 
Let's see a few examples: 

* Windows: `setwd("C:/Users/Documents/SNA4DSworkingFiles")`
* Mac: `setwd("/Users/Documents/SNA4DSworkingFiles")`
* Linux: it depends, but most likely the Mac way

The way to know for sure is by clicking on the folder you want to set up as a 
directory and take its 'address' from 'Properties.' Then copy and paste the 
location, but make sure to substitute \\ with /. 

### .txt and .csv

A text file (.txt) is a kind of computer file structured as a sequence of lines 
of electronic text. Sometimes it is also spelled textfile; 
an old alternative name is flatfile.

A comma-separated values (.csv) file is a delimited text file that uses a comma
to separate values. Each line of the file is a data record. Each record consists
of one or more fields, separated by commas. Sometimes instead of commas, you 
have semicolons.

*R* has an inbuilt function to import *.txt* data and turn it into a data frame:

`read.table()`

*R* also has an inbuilt function to import *.csv* data and turn it into a data 
frame:

`read.csv()`

If you use the help function `?` with either of these two functions, you will be
sent to the same help file since these two functions work in the same way. 
Several arguments in both functions will help you to customize your data frame.
Let's see the essential arguments together. 

* *file* = you pass the file's name within quotation marks, including the file 
extension. For example `read.csv(file = "data.csv")`. In *R*, you can also omit 
the argument's name and write `read.csv("data.csv")`.

* *header* = if you set the header as TRUE, the first row in your imported data 
will be considered column names in the data frame you create. In case your 
columns don't have a name yet, set it as false. Otherwise, your first observation
won't be considered as such anymore. 

* *sep* = this is the separator. It is either "," or ";" for *csv* data and " " 
empty for *txt* data.

* *stringsAsFactors* = if your data contains character variables, *R* will read 
them as characters unless you specify that they are factors setting this 
argument to TRUE.

* *na.strings* = if you know that you are importing a file with missing data, 
you should use this argument to help *R* to recognize them. So if the missing 
data in your file is expressed as NA, you can use this argument as
`na.strings = "NA"`. 
However, let's say that your missing data is expressed by the word *NO* and by
this sign *..*, you should specify `na.strings = c("NO", "..")`.


let's consider this data file called *data.csv*, 

```{r echo = FALSE, results = 'asis'}

cn <- c("age", "gender", "eyes_col")
one <- c(7, "M", "BLUE")
two <- c(8, "F", "BROWN")
three <- c(8, ".", "GREEN")
four <- c(7, "F", ".")

tab <- rbind(cn, one, two, three, four)
rownames(tab) <- NULL
knitr::kable(tab)
```


Hence you would import it into *R* this way

`data <- read.csv('data.csv', sep = ",", header = TRUE, stringsAsFactors = TRUE,
                na.strings = ".")`



### Excel
Everyone nowadays is familiar with Microsoft Office and with its spreadsheet 
editor called Excel. And you surely already know that excel has its own 
extension... Actually, several of them, since they evolved over time. It is
possible to import 
excel data into *R*. 

If it is a simple spreadsheet, the easiest way is to save it as a *.csv* or 
*.txt* 
file and import it that way. Otherwise, you need to use an *R* package that 
can do the job for you.

According to which extension (how old is your file), you need different packages. 
We will cover together the extension *.xlsx*.

You should install the package `readxl` by running the code:

`install.packages("readxl")`

After that, you will be able to use the function `read_xlsx()`. Importing data 
in this way gives you the possibility of selecting specific columns and rows 
from one specific spreadsheet. 

For example:

`mydata <- readxl::read_xlsx("excel_data.xlsx", sheet = 2 , range = "A2:U38")`

Will import into *R* the data included in the range column A row 2 and column U row 38 taken from the second spreadsheet in your excel file.


### R data Format

After you worked for a while on your data, you want to save what you already did 
into a new data object (and eventually open it again!)

R has its own systems to save data. Let's explore two of them.

#### RDS

If you want to save one object to a file you can use the function `saveRDS()`

For example, `saveRDS(object, file = "my_data.rds")`

where *object* is the name of the variable storing your data and *my_data.rds* 
will be the name of the file saved in your directory.

When you will get to the point of re-using the object you stored (good for you!)

Restore the object with `readRDS()`. 

For instance, `readRDS(file = "my_data.rds")`


#### RDATA

If you want to save one OR MORE objects to a file you can use the function 
`save()` (how creative!)

For one object:
`save(data1, file = "data.RData")`

or to save multiple objects
`save(data1, data2, file = "data.RData")`


To load the data again in both cases you can use
`load("data.RData")`

who would have expected such an original name ;)


## Inspecting data

After you loaded your data, unless it is very small and you can visually inspect 
it, you will realize that you have NO IDEA about what is inside that blurb you 
just imported into *R*. Instead of panicking (ok, you can panic for 10 seconds), 
use the functions that *R* already set up for you to explore your data! 
Let's take a look at our options.


### print

If you want to inspect your data visually, you should simply type the object 
name into the console. Let's use a demo dataset called *mtcars*. 
Guess what it is about? cars :)


Just type the name there and see what happens!

```{r try_something251, exercise = TRUE, echo = TRUE}
mtcars
```

Ok, it worked, but we still don't know much about it, right? 

Shall we try something else?


### A glimpse inside

One of the problems with printing the data frame is that when there are many 
rows, it is not helpful to print them all. We can use `head()` or `tail()` 
instead to overview the first or last six rows in the data frame. 

Give it a go!

```{r try_something25, exercise = TRUE, echo = TRUE}
head(mtcars)
tail(mtcars)
```


It might also be helpful to know how big it is our data frame. We can do that 
with `nrow()` and `ncol`. Guess which one does what? :)

```{r try_something26, exercise = TRUE, echo = TRUE}
nrow(mtcars)
ncol(mtcars)
```
Still, don't we want something more?

### summary

The function `summary()` is our chance to get a SUMMARY... (who would have 
thought? LOL) Why don't you try? 

```{r try_something27, exercise = TRUE, echo = TRUE}
summary(mtcars)
```

Much better than the previous options, right? 
For each variable in the data frame, we get information that helps us know the data. 

This data is numeric only. Do you think it will work with other kinds of data 
though? why don't you run a summary with the dataset *iris*?

```{r try_something28, exercise = TRUE, echo = TRUE}
summary(iris)
```

*iris* is about flowers in case you haven't notices :)

### str

Another helpful function is `str()`. It compactly displays the structure of an
*R* Object. 

Try with one of the data set above.

```{r try_something29, exercise = TRUE, echo = TRUE}
str(iris)
```

Isn't it very informative? 



## Bits of Data Manipulation

Manipulating data isn't something that will lead you in front of the European 
court of human rights. Data aren't people, and we can manipulate them as much 
as we want, but only
if we know how to do it. 

Otherwise, it is more likely that the data will manipulate us, and we won't 
understand the story they hide! we don't want this to happen, 
hence let's learn a few tricks! 

### R counts elements from 1 

Some languages (Python) count the elements from 0. 

In those languages (Python), if you have a vector with three elements, they will 
be element 0, element 1, and element 2. 

*R* does not believe in this kind of zeros (in other it does), and it counts 
from 1, so if you have a vector with three elements, they will 
be element 1, element 2, and element 3. 

This is a crucial piece of information if you want to fish those elements from 
any data structure. 

### Fishing from vectors

If I give you a vector, and I suggest to make use of a particular kind of net
(square brackets) can you fish the third element for me? :)



```{r grade_code121, exercise = TRUE, echo = TRUE}

x <- c(674, 4186, 5308, 5083, 6140, 6381)

```


```{r grade_code121-solution}

x <- c(674, 4186, 5308, 5083, 6140, 6381)
x[3]

```

```{r grade_code121-check}
gradethis::grade_code(correct = "Well done!")
```

As a prize I can also tell you that you can fish more efficiently: why take one 
element per time when you can do more? I know... consumerism...

```{r try_something30, exercise = TRUE, echo = TRUE}

x <- c(674, 4186, 5308, 5083, 6140, 6381)
x[2:4]
x[c(1, 3, 4)]

```

Finally, Can you replace element two with the number '123'? 
Hint: assign the new number on top of the old one :)

```{r grade_code12, exercise = TRUE, echo = TRUE}

x <- c(674, 4186, 5308, 5083, 6140, 6381)


```


```{r grade_code12-solution}

x <- c(674, 4186, 5308, 5083, 6140, 6381)
x[2] <- 123

```

```{r grade_code12-check}
gradethis::grade_code(correct = "Yes!")
```



### Fishing from data frames and matrices

If you need to take an element from a data frame or a matrix, you can use 
square brackets [ , ]

The space on the left-hand side of the comma represents rows and the space on 
the right-hand side columns. 

let's take the first column from this data frame


```{r try_something31, exercise = TRUE, echo = TRUE}

df <- data.frame(col1 = 'one', col2 = 1:10)

(c <- df[ , 1])

```


Can you take for me the third element in the second row from this matrix below, 
assign it to the variable *e* and print it?


```{r grade_code15, exercise = TRUE, echo = TRUE}

m <- matrix(1:9, 3, 3)

```


```{r grade_code15-solution}

m <- matrix(1:9, 3, 3)

(e <- m[2, 3])

```

```{r grade_code15-check}
gradethis::grade_code(correct = "Here you go!")
```


When you are working with data frames, you also have the option to use the 
dollar sign $. Let's take the third element from column two using the dollar sign.

```{r try_something32, exercise = TRUE, echo = TRUE}

df <- data.frame(col1 = 'one', col2 = 1:10)

(c <- df$col2[3])

```

easy right?

### Fishing from lists. 

When you fish from lists, you need to consider that you are fishing on two 
levels. First from the list itself, second from the data structure you have 
inside stored as an element. 

We have a list here below, and we don't know what is in there. Let's access the 
list level only and use `class()` to check what to do next. 

```{r try_something33, exercise = TRUE, echo = TRUE, exercise.setup = "try_something24"}

L <- list(A, B, C)

L[1]
class(L[1]) 

L[[1]]
class(L[[1]])


```

If you use one square bracket only, it will print the content, but the class is 
still *list*.

If you use double square brackets, it will still print the content, but it will 
also, get to the class of the content. 

Can you take column 1 in the data frame stored as the first element in the list?


```{r grade_code16, exercise = TRUE, echo = TRUE, exercise.setup = "try_something24"}

L <- list(A, B, C)

```


```{r grade_code16-solution}

L <- list(A, B, C)

L[[1]]$col1

```

```{r grade_code16-check}
gradethis::grade_code(correct = "Fantastic!")
```



### Subsetting 
Very often, you import some data, but you don't need everything. It might be cumbersome and confusing to keep variables that are unnecessary for the analysis you are carrying out. Hence you need to subset your data and keep only the portion that is helpful for your goals. 

There are several ways to do it, and also several packages to make this work easier. In this tutorial, we are only going to cover the options we have in *R-base*.

Mainly using the data manipulation tricks we just learned and the function 
`subset()`. 

Can you please subset for me the columns `mpg`, `cyl` and `disp` from the data 
set *mtcars* using square brackets? 

```{r grade_code171, exercise = TRUE, echo = TRUE}
str(mtcars)

# mtsub <- 

```


```{r grade_code171-solution}

str(mtcars)

mtsub <- mtcars[ , 1:3]
```

```{r grade_code171-check}
gradethis::grade_code(correct = "Fantastic!")
```


How about deleting these three columns instead? 
Hint: Use the operator - and round brackets...

```{r grade_code172, exercise = TRUE, echo = TRUE}
str(mtcars)

# mtsub <- 

```


```{r grade_code172-solution}

str(mtcars)

mtsub <- mtcars[ , -(1:3)]
```

```{r grade_code172-check}
gradethis::grade_code(correct = "Awesome!")
```

How would you subset a data frame in order to keep only certain values? 

Let's say that you want to keep only cars with `mpg` larger or equal to 20. 
Since the `mpg` values are provided in the rows, you need to pass the 
information on the left-hand side of the square brackets [x, ]. Hence you tell
*R* that you are interested in the `mpg` column, specifically in those values 
only.

Run this code and inspect the result.


```{r try_something34, exercise = TRUE, echo = TRUE}

(mtsub <- mtcars[mtcars$mpg >= 20 , ])

```


Did you get it? 

Let's see if you can subset for me the cars with 4 gears. 


```{r grade_code17, exercise = TRUE, echo = TRUE}
str(mtcars)

# mtsub <- 

```


```{r grade_code17-solution}

str(mtcars)

(mtsub <- mtcars[mtcars$gear == 4 , ])
```

```{r grade_code17-check}
gradethis::grade_code(correct = "YES!")
```


Using square brackets can get time consuming if you want to do several subsetting 
operations at the same time. For those cases we have the `subset()` function.

```{r try_something351, exercise = TRUE, echo = TRUE}

mtsub <- subset(mtcars, mpg <= 20, select = c("mpg", "gear", "wt"))


```

That makes it faster, doesn't it? 

Why don't you inspect the `subset()` function using the help `?`, and subset for
me the entire data set excluding one column `carb` and excluding all the values 
of `disp` below 100?


```{r grade_code18, exercise = TRUE, echo = TRUE}
str(mtcars)

# mtsub <- 

```


```{r grade_code18-solution}

str(mtcars)

(mtsub <- subset(mtcars, -disp < 100, select = -carb ))

```

```{r grade_code18-check}
gradethis::grade_code(correct = "Well done!")
```


### Recoding

When you do data manipulation, very often, you need to replace some value with 
another. For example, you can classify cars according to how many gears they have. 

Let's try using the `table()` function to see how many gears we have. 


```{r try_something35, exercise = TRUE, echo = TRUE}

table(mtcars$gear)


```


Ok, we have cars with 3, 4, or 5 gears. Let's say that I like to speed, and for 
me, cars with only 3 or 4 gears are not cool at all. 

Let me recode it this way.

```{r try_something36, exercise = TRUE, echo = TRUE}

# create a new col gearR

mtcars$gear[mtcars$gear == 3] <- "not_cool"
mtcars$gear[mtcars$gear == 4] <- "not_cool"
mtcars$gear[mtcars$gear == 5] <- "cool"

table(mtcars$gear)
```

Done! There are only five cars that are cool enough for me ^_^.


```{r remov, echo = FALSE, include = FALSE}
rm(mtcars)
```

Anyway, this square brackets recoding can easily get to the nightmare level if 
you have too many things to recode. A MUCH better way to do it is to use 
`ifelse()`. 

Do you remember the conditions we learned about before in this tutorial? Well, 
let's put them to good use. 

```{r try_something37, exercise = TRUE, echo = TRUE}

# create a new col gearR
mtcars$gear <- ifelse(mtcars$gear == 5, "cool", "not_cool")

table(mtcars$gear)
```

How much better is this?? Let's understand what happened. 

`ifelse()` has three conditions:

`ifelse(1, 2, 3)` 

You read it as: If 1 is TRUE, do 2, else do 3. 

Hence, in our example: If the gear is equal to 5, recode 5 with 'cool'; in any 
other case, recode it with 'not_cool'. 


Why don't you try to recode `cyl`, put the highest number as 'cool', and all the 
rest as 'not_cool'? 

```{r grade_code19, exercise = TRUE, echo = TRUE}

table(mtcars$cyl)

 

```


```{r grade_code19-solution}


table(mtcars$cyl)
mtcars$cyl <- ifelse(mtcars$gear == 8, "cool", "not_cool")

```

```{r grade_code19-check}
gradethis::grade_code(correct = "Well done!")
```


Take into account that you can nest `ifelse()` into each other and set up as many 
conditions as needed!

### The Pipe

The very last version of *R* included a helpful tool called the pipe. The pipe, 
represented by this symbol `|>`, allows you to pipe a value into a call 
expression or a function expression.

A pipe expression passes, or pipes, the result of the left-hand side expression 
to the right-hand side expression. This is a bit revolutionary for the *R* way 
of thinking, and it allows you to perform certain kinds of manipulation in a 
very quick way. 

Let's try to subset flowers belonging to the species 'setosa' only with the pipe.

```{r try_something381, exercise = TRUE, echo = TRUE}
iris |>
   subset(Species == "setosa")
```

That was quick, right? 

But you might want to apply a function to the entire data set; let's say you 
need to see the structure

```{r try_something382, exercise = TRUE, echo = TRUE}
iris |>
   str()
```

For a short expression like this, it does not make much difference than 
`str(iris)`. 
Still, in other situations, it might be handy. If you compare these two 
different ways of applying a function to a data set, you can understand 
why the pipe is a revolutionary way of thinking for the *R* system. 
Honestly? doesn't it remind you of that other slithering programming languages 
we mentioned before? :) 


## R-base VS packages

"Everything you can do in *R-base* must be done in *R-base*. 
For all the rest, use packages." 

Just imagine Yoda saying that! :) 

Fair enough, right? Still, let's see how to use packages, shall we? 

### Installing, Removing, Updating

In *R* packages are also called libraries. You can install them with the 
function `install.packages()` if they are available in the Comprehensive R 
Archive Network, better know as CRAN. It is a different story if they are 
available on GitHub or other similar platforms, and we won't cover it here. 

Hence, to install one package from CRAN, you will run

`install.packages("package_name")`

you can also install more than one package at the same time


`install.packages(c("package_name1", "package_name2"))`


If you then need to remove a package (for whatever reason), you can run


`remove.packages("package_name")`

This last function will also work for packages coming from different sources 
than CRAN.

In order to update CRAN packages, there is the convenient function

`update.packages("package_name")"


###  Using Packages
After installing it, a package is not immediately available for you to use, but 
you have to "call" it. 

You can do that in several ways, but the safest is to call the package name, 
followed by :: (colon colon) and function name. 

Such as: package_name::function(whatever the function contains)

this can also be done with *R-base*, but it is unnecessary since *R-base* 
functions are always available. For example we can call the `table()` function 
as: 

```{r try_something38, exercise = TRUE, echo = TRUE}

base::table(mtcars$hp)

```

This is the safest way since several packages use the same names for their 
functions, and you might end up calling a function and not knowing what it 
does since it has two or more different versions in two or more different 
packages. 
 

Also, some packages conflict with each other, and your *R* session might even 
crash due to this. 

There are situations in which you need to attach a package, such as when you are
doing certain visualizations (not all of them). You should attach a package ONLY
in those circumstances where you are forced to do it. 

You can do it in two ways: 

`library(package_name)` with or without quotation marks, it doesn't matter. 

OR

`require(package_name)`


in both cases, after you are done with the package, you MUST detach it by 
running: 

`detach("package:package_name", unload=TRUE)`

Only by detaching it can you avoid potential conflicts with other packages. 



## Functions 
So far in this tutorial, we used plenty of functions. Still, we haven't written 
one yet! Even if there are billions of functions out there, there are also 
gazillions of problems to solve, and sometimes you need to be able to step up 
and write your own solution to a computing problem.

Let's fix this!


### Defining a function

Functions are defined by code with a specific format:

They always have

* One function name
* Arguments
* Some operation inside (otherwise, why would we be here?)
* return the value(s)

For example:

`function.name <- function(arg1, arg2, arg3, ...) {

  newVar <- (sqrt(arg1) + sqrt(arg2)) / arg3  # do Some Useful Stuff
  newVar    # return value 
  
}`


In *R* a function ALWAYS uses {curly braces} :) 

Would you please build a function for me?
I need one that has 4 arguments. Take the sum of arg1 and arg2, divide them by 
the summation of arg3 and 4. Call the function 'helper' 
(cause you are helping me :) !)

```{r grade_code20, exercise = TRUE, echo = TRUE}



 
```


```{r grade_code20-solution}

helper <- function(arg1, arg2, arg3, arg4, ...) {

  newVar <- (arg1 + arg2) / (arg3 + arg4)  
  newVar    
  
}

```

```{r grade_code20-check}
gradethis::grade_code(correct = "Congrats!")
```



### Calling functions

Thanks for the `helper()` function you made for me! Shall we try it? 



```{r try_something39, exercise = TRUE, echo = FALSE, include = FALSE}
helper <- function(arg1, arg2, arg3, arg4, ...) {

  newVar <- (arg1 + arg2) / (arg3 + arg4)  
  newVar    
  
}
```

Consider the *iris* data set again and employ the function in this way

* arg1 <- Sepal.Length
* arg2 <- Sepal.Width
* arg3 <- Petal.Length
* arg4 <- Petal.Width

assign the result to the `f` variable


```{r grade_code21, exercise = TRUE, echo = TRUE, exercise.setup = "try_something39"}

# str(iris)

 
```


```{r grade_code21-solution}

# str(iris)

f <- helper(iris$Sepal.Length, iris$Sepal.Width, iris$Petal.Length, iris$Petal.Width)



```

```{r grade_code21-check}
gradethis::grade_code(correct = "It's lovely!")
```



Ideally you should save your function in a clean *R* script and import it for 
use by running

`source("mypath/myfun.R")`



## Loops!

As you very likely know already, loops are tools that help you in making a 
repetitive task several times. All modern programming languages provide special constructs that allow for the repetition of instructions or blocks of instructions.

There are two main types of loops:


* while loops 

* for loops

Let's get to know them.

### While loops

Do you remember the conditions we already used a lot? Well, a while loop does
something for you *while* a condition is TRUE. How nice of it? 

This is the logic

while (test_expression)

{

statement

}

The while loop will stop doing whatever it is doing when the condition expires. 
For instance, you eat a yogurt jar while it is good. When it expires, you stop 
eating it and toss it away (Well, at least we don't eat when mold is inside...YAK!!) 

Let's try with a simple example.


```{r try_something40, exercise = TRUE}

i <- 1

while (i < 10) {
  
cat(i)
i = i+1

}

```

In this example, the while loop is printing (the function `cat()`) a progression
of numbers, but only while they are still smaller than 10. 



### For Loops

Do you remember when Bart Simpson had two write the same sentence over and over? 
Bart would have needed a for loop! 

The for loop does the same thing as many times as you tell it to do it! 
Lovely, isn't it?

The logic is

for (this number of times that I'm telling you here)

{

Enter the data structure and do this for me!

}

Let's say that you want to print the sentence *The year is ...* for as many 
years as you need, that's how you do it!


```{r try_something411, exercise = TRUE}

y <- c(2010, 2011, 2012, 2013, 2014, 2015)

for (i in 1:length(y) ){
  
  print(paste("The year is", y[i]))
  
}

```

Let's break it into pieces:

* `y` is the vector that provides the year, hence the dimension. 
* `i` is our index. It is our variable that enters the data structure. 
It could be anything 
* 1:`length(y)` tells the loop how many times it needs to repeat the operation.
* `print()` prints (who would have thought?!)
* `paste()` glues together the character string and the information that the index
retrieves from the data structure. 

Clear? 

Now,  you try!

You have a vector of 6 Kids and a vector of 3 Animals. If your indexes are respectively *kid* and *animal*, can you write a nested for loop so that you print a total of 6*3 sentences where each kid loves each animal? For example:

"Anna loves cats"

The structure of a nested loop is

for(...) {

	for(...) {

  statement

	}
	
}

It's a little difficult, but I KNOW THAT YOU CAN DO IT!!! Cal your indexes *kid*
and *animal*. As *kid* in *Kids* and *animal* in *Animals*.


```{r grade_code23, exercise = TRUE, echo = TRUE}

Kids <- c("Anna", "Barbara", "Max", "Luke", "Jack", "Martha")
Animals <- c("cats", "dogs", "crocodiles")

 
```


```{r grade_code23-solution}

Kids <- as.factor(c("Anna", "Barbara", "Max", "Luke", "Jack", "Martha"))
Animals <- c("cats", "dogs", "crocodiles")

for (kid in 1:length(Kids) ){
  
  for(animal in 1: length(Animals)){
    
    print(paste(Kids[kid], "loves", Animals[animal]))
    
  }
}


```

```{r grade_code23-check}
gradethis::grade_code(correct = "You are on fire!")
```


### For loops with conditions

For loops can also be used in more elaborate ways. For instance, we can recode 
a variable using a for loop. 

Bear with me! We have information about two kids, and we want to compare them. 
So we want to create the variable *comp* that stands for comparison. 
If they have the same feature, we will read "yes", otherwise "no". 

To do that, we initialize the variable *comp*, and we make use of a temporary 
variable *temp* to store the result. 

Let's do it!

```{r try_something412, exercise = TRUE}

mary <- c('blue', 8, 'F', 'Dutch')
tom <- c('green', 8, 'M', 'Dutch')
df <- data.frame(mary, tom)
rownames(df) <- c('eyes', 'age', 'gender', 'nationality')
df

comp <- NULL

for (i in 1:nrow(df)) {
  
  if(df[i, 1] == df[i, 2] ) {
    
    temp <- 'yes'
  
  } else {
    
    temp <- 'no'
    
  }
  
  comp <- append(comp, temp)
}

df <- cbind(df, comp)

df
```

Now we have the variable *comp* in the data frame! In such a small data frame, 
it does not seem so useful, but it can make a real difference in a large one! Also, 
the variable *comb* is ready for you to do some other data manipulation!


Your turn to play!
Can you tell me who won the football matches? 
You have a list of four football matches. We stand for team 1. The score on the 
left always stands for team 1 and the one on the right for team 2. Can you make 
*R* print "Win" when team 1 wins and "Lose" when team 1 loses? 

call your index *i*

```{r grade_code24, exercise = TRUE, echo = TRUE, exercise.setup = "try_something39"}

matches <- list(c(2, 1), c(2, 5), c(6, 3), c(1, 2))

 
```


```{r grade_code24-solution}

matches <- list(c(2, 1), c(2, 5), c(6, 3), c(1, 2))

for (i in matches){
  
    if (i[[1]] > i[[2]]){
      
        print("Win")
      
    } else {
      
        print ("Lose")
    }
}


```

```{r grade_code24-check}
gradethis::grade_code(correct = "Yeahhh!")
```


## The Apply family

As we have just seen, *R* uses for loops as much as any other programming language. However, in *R*, there is an alternative to perform the same task. This 
is the Apply family! 

The apply family consists of vectorized functions that minimize your need to create loops explicitly. Vectorization is the operation of converting repeated operations on simple numbers (“scalars”) into single operations on vectors or matrices. A collection of numbers is a numeric vector. If you combine vectors 
(of the same length), you obtain a matrix. You can do this vertically or horizontally, using different R instructions. Many for loops can be made implicit by using vectorization, and this operation makes the computation more efficient.

The apply family is large! This is an overview:

* `apply()` for matrices and data frames
* `lapply()` for lists…output as list
* `sapply()` for lists…output simplified
* `tapply()` for vectors
* `mapply()` is a multivariate version of sapply.
* `rapply()` is a recursive version of lapply
* `vapply()` vapply is similar to sapply, but has a pre-specified type of return 
value

We are going to explore `apply()` together, and you can figure the others out by 
yourself.

### Apply

The `apply()` function is most often applied to the rows or columns (margins) of matrices or data frames. It can also be used for arrays.

`apply()` has three arguments.

* X = the matrix or data frame we want to process
* MARGIN = a vector telling the function what to process
* FUN = the function you want to apply to the data structure

Why don't we try this out? 

We consider the numeric columns of *iris* hence `iris[1:4`. Since we are focusing on columns, the second argument is `2`. We pass to `apply()` the function `mean()`

```{r try_something41, exercise = TRUE}

apply(iris[1:4], 2, mean)
```


Done! We have the mean of each column! Easy peasy!



Can you estimate the sum by row considering only the first two columns? 

```{r grade_code25, exercise = TRUE, echo = TRUE}


 
```


```{r grade_code25-solution}


apply(iris[1:2], 1, sum)

```

```{r grade_code25-check}
gradethis::grade_code(correct = "Easy right?!")
```



You can also combine several functions in one go. This example is taking the 
`min()` and the `max()` value of each of the four columns and estimating the 
`mean()` between these two values for each column. 

```{r try_something42, exercise = TRUE}
apply(iris[1:4], 2, function(x) mean(c(max(x), min(x))))
```



You can be creative with the `apply()` family as much as with for loops!


## Very basic r graphics

Doing good visualizations is almost an art. However, everyone, with a bit of 
practice, can produce images to show data.

We visualize data for several reasons, but one of the most 
important is to make data and results clear and understandable for a 
larger public. Communicability is an essential skill both in academia 
and in the private sector. 

You might have very important findings, but if your audience does not
understand them, they are worth nothing. 
A great visualization makes more than half of the job! 
Great visualizations are story tellers!

There are several packages that allow you to visualize. For now, you should get 
familiar with the *R-base* way of doing. 



### Scatter plots

A scatter plot (also called a scatterplot, scatter graph, scatter chart, 
scattergram, or scatter diagram) is a type of plot or mathematical diagram using 
Cartesian coordinates to display values for typically two variables for a set of 
data. 

The most important function for making visualization in *R-base* and for making 
scatterplots is `plot()` (weird choice, right :)?). 
Let's generate some data and play with it. 

In the code box below, there is a `plot()` function with 9 arguments. What are 
are they doing? 

* `x` = x axis 
* `y` = y axis. In the plot below is set to `y1`, set it to `y2`. What is the difference?
* `type` = it tells the plot 'how' to visualize the data. In the example below is set to "p", and it prints points. Observe what happens if you set it to "b" and then to "h".
* `pch` = tells the function to add points to your plot and how to plot them. 
Here is set to 16, but there are many options.
* `frame` = it tells the function to frame your plot. Could you set it to FALSE and see what happens?
* `col` = chooses the color of your points. Now is set to "red", but there is an entire palette of colors to explore. Try to set it to "#2E9FDF" instead.  
* `xlab` = puts a label on the x axis
* `ylab` = puts a label on the y axis
* `main` = prints a title on the top of your plot

```{r try_something43, exercise = TRUE}
x <- 1:10
y1 <- x*x
y2  <- 2*y1

plot(x = x, 
     y = y1, 
     type = "p", 
     pch = 16, 
     frame = TRUE,
     col = "red", 
     xlab = "x", 
     ylab = "y", 
     main = "My Plot")
```


It's not too hard! Right? Now it's your turn. 

Make a plot using the *iris* data. Place *Sepal.Length* on the x axis, and 
*Sepal.Width* on the y one. Make it print points, choosing option 16. Make it 
blue. Put x and y labels with the names of the variables, and set 
"Iris Sepal Plot" as the main title.


````{r grade_code261, exercise = TRUE, echo = TRUE}


 
```


```{r grade_code261-solution}


plot(x = iris$Sepal.Length, 
     y = iris$Sepal.Width,
     type = "p",
     pch = 16, 
     frame = FALSE,
     xlab = "Sepal.Length", 
     ylab = "Sepal.Width", 
     col = "#2E9FDF", 
     main = "Iris Sepal Plot")

```

```{r grade_code261-check}
gradethis::grade_code(correct = "Easy right?!")
```



There are very many parameters that you can customize to make your plot fancy and pretty. When you will need to plot for real, spend some time finding 
solutions that work for your case! the help function gives you access to all 
the knowledge you need!

Let' take a look at other kinds of plots now!

### Box Plots

In descriptive statistics, a box plot or boxplot is a method for graphically depicting groups of numerical data through their quartiles. Box plots may also have lines extending from the boxes (whiskers), indicating variability outside the upper and lower quartiles; hence the terms box-and-whisker plot and box-and-whisker diagram. 

The function to make a boxplot in *R-base* is `boxplot()` :)

You can have a boxplot displaying one variable's mean and standard deviation.

```{r try_something44, exercise = TRUE}

boxplot(iris$Sepal.Length)

```

Or a boxplot displaying that variable by categories. 

Comment on `notch` and see the difference. 

Try to change the border color.
```{r try_something45, exercise = TRUE}

boxplot(Sepal.Length ~ Species, 
        data = iris, 
        frame = FALSE, 
        notch = TRUE, 
        col = c("#999999", "#E69F00", "#56B4E9"),
        border = c("black", "black", "black"))

```


Note that in this example the second chosen variable is a factor.

Also, boxplot can be made very fancy with some time and enthusiasm!


### Bar Charts and Histograms 

Bar charts and Histograms might look alike, but they are conceptually very different. 

A bar chart or bar graph is a chart or graph that presents categorical data with 
rectangular bars with heights or lengths proportional to the values that they 
represent. The bars can be plotted vertically or horizontally.

A histogram is an approximate representation of the distribution of numerical 
data. 

Histograms are used to show distributions of variables while bar charts are used
to compare variables. 

Histograms plot quantitative data with ranges of the data grouped into bins or 
intervals while bar charts plot categorical data. 

The function we need this time are `barplot()` and `hist()`

Let's get to bar charts!

Bart chart represent categorical data, right? Let's use a new 
data set that inform us about death rates by age. 

Let's explore the new data set and plot the death rate of rural male by age.

```{r try_something46, exercise = TRUE}

str(VADeaths)

barplot(VADeaths[ , "Rural Male"], 
        xlab = "Rural Male", 
        ylab = "rate",
        main = "Rural Male's Death by Age")

```

Let's get horizontal!

```{r try_something47, exercise = TRUE}

barplot(VADeaths[ , "Urban Female"], 
        horiz = TRUE,
        xlab = "Urban Female", 
        ylab = "rate",
        main = "Urban Female's Death by Age")

```


Let's get to color!


```{r try_something48, exercise = TRUE}

barplot(VADeaths[ , "Urban Female"], 
        xlab = "Urban Female", 
        ylab = "rate",
        main = "Urban Female's Death by Age",
        col = "white",
        border = c("blue", "cyan", "green", "pink", "red"))

```

Can you make a bar chart of the variable "Rural Female", where each bar has a 
different color, and every bar has a black border? 

````{r grade_code26, exercise = TRUE, echo = TRUE}


```


```{r grade_code26-solution}

barplot(VADeaths[ , "Rural Female"], 
        xlab = "Rural Female", 
        ylab = "rate",
        main = "Rural Female's Death by Age",
        col = c("blue", "cyan", "green", "pink", "red"),
        border = "black")
 
```

```{r grade_code26-check}
gradethis::grade_code(correct = "Good Job!")
```




Let's get to Histograms. They plot quantitative data with ranges of the data 
grouped into bins or intervals

Shall we try then? Let's use the numeric variable *mpg* from *mtcars*


```{r try_something51, exercise = TRUE}

hist(mtcars$mpg)


```

ok, the plot runs. What does this mean? We can use the function table to understand
what it is going on here. 

```{r try_something53, exercise = TRUE}

hist(mtcars$mpg)
table(mtcars$mpg)


```

From table, we know that the min value is 10.4 and the max value is 33.9. We can 
see this range in the plot. 

The function decided to cluster all the observations into six clusters. 
So we know that there are six observations in between 10 and 15. Check the frequencies 
produced by the table function. It is correct, right? 

Ok, but we don't like these six groups, or bins. No problem. We can change the 
granularity with the argument `breaks`. 

The range between the max and the mean is roughly 34 - 10 = 24. Let's try to break
into 24 bins.




```{r try_something52, exercise = TRUE}

hist(mtcars$mpg, 
     breaks = 24)

```

Better now? Does it make more sense this way? :) I hope so. 

This way you can see the distribution of the variable... Let's say a bit more 
about it.



### Density Plot

Do you know what is a distribution? 

In probability theory and statistics, a probability distribution is the 
mathematical function that gives the probabilities of occurrence of different 
possible outcomes for an experiment.


Let's say that I have a dice, and I toss it for 10 times. Vector `x` stores the 
result I get each time. It can range from 1 to 6. 

If we want to observe the probability of occurrence of each of these 10 outcomes 
we can estimate and plot the *Kernel density* of the variable `x`

```{r try_something49, exercise = TRUE}

x <- c(1, 2, 3,  3, 3, 4, 4, 5, 5, 6)

density(x)

plot(density(x))
```


Do you remember that stat class when you heard about normal distributions? Well, 

`x` has almost a normal distribution.

Let's plot it in a nicer way.

```{r try_something50, exercise = TRUE,  exercise.setup = "try_something49" }


dens <- density(x)


plot(dens, frame = FALSE, col = "steelblue", 
     main = "Density plot of x") 


```

Or even a nicer option

```{r try_something55, exercise = TRUE,  exercise.setup = "try_something50" }

plot(dens, frame = FALSE, col = "steelblue", 
     main = "Density plot of x") 

polygon(dens, col = "steelblue")
```




## The End

That's all for now! Hopefully you learned something helpful and did not get too 
bored! See you in the next tutorial!
